---
title: "Basic popgen of plurivora"
subtitle: "Populations based on Town and Source"
author: "Nick Carleson"
date: "November 14, 2017"
tags: [Town, Source]
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Populations
### How do the _Phytophthora pluriva_ populations cluster?

We'll need to attach the metadata and get a hierarchy going for the 69 samples that survived filtering. Then we'll make PCA, structure, and Fst plots.

This picks up after depth_missingness_coding_genotypedgvcf.Rmd

## Data review
### Load files
```{r load, results="hide", message=FALSE}
library(vcfR)
library(adegenet)
library(ape)
library(poppr)
library(dplyr)
library(treemap)
library(ade4)
library(magrittr)
library(hierfstat)
library(gplots)
library(reshape2)
library(ggplot2)

load("~/Data/plurivora/pst1_vcf3")
load("~/Data/plurivora/pst1_genind")
load("~/Data/plurivora/pst1_genlight")
```

### Initializing objects
Check out the objects and prepare them
```{r inspect}
pst1_vcf3

# Convert from genind to genclone for further processing
pst1_genclone <- as.genclone(pst1_genind)

# Load in definitions of strata
strata_defs <- read.table("~/Documents/plurivora/GBS_plurivora_strata_for_uneak.csv",
                          sep = ",", header = TRUE, stringsAsFactors = FALSE)
# Reduce the number of defined individuals in key to only filtered individuals
# for genclone
strata_defs_filtered <- subset(strata_defs[,], strata_defs$new_name %in% rownames(pst1_genclone@tab))
strata(pst1_genclone) <- strata_defs_filtered
# for genlight
strata_defs_filtered <- subset(strata_defs[,], strata_defs$new_name %in% pst1_genlight@ind.names)
strata(pst1_genlight) <- strata_defs_filtered
```

### Setting populations
Now get ready to begin some analysis and do first steps
```{r begin analysis}
# Set populations based on strata
setPop(pst1_genclone) <- ~Town/County
setPop(pst1_genlight) <- ~Town/County
pst1_genclone
# for genclone
poppr(pst1_genclone)
# for genlight
diversity_stats(pst1_genlight)

# Reassigning strata
pststrata <- strata(pst1_genclone) %>%
  group_by(Town, County, Source, Nursery) %>%
  summarize(Count = n())
pststrata
```

### Visualize what's in our data
This will get us a first look at what composes our dataset
```{r first viz}
# Adjusting the aesthetics for the labels
label_position <- list(c("center", "top"), c("center", "center"), c("center", "bottom"))
label_size     <- c(Town = 15, County = 10, Source = 10, Nursery = 0)

# Plotting, First three arguments are necessary.
treemap(dtf = pststrata, index = c("Town", "County", "Source", "Nursery"), vSize = "Count",
        fontsize.labels = label_size, align.labels = label_position, vColor = "Town",
        title = "P. plurivora GBS samples")

```

### Clone correction and Locus summaries
Does clone correction have an impact?
```{r clones}
mcc_TY <- clonecorrect(pst1_genclone, strata = ~Town/County, keep = 1:2)
# Before clone correction
pst1_genclone
# After clone correction
mcc_TY
# apparently no difference...moving on anyway.
cc <- locus_table(mcc_TY, info = FALSE)
mp <- locus_table(pst1_genclone, info = FALSE)
locus_diff = mp - cc
barplot(locus_diff[, "1-D"], ylab = "Change in Simpson's Index", xlab = "Locus",
main = "Comparison of clone-corrected vs. uncorrected data")
# There is no difference at all. SUPER weird. Looking at the locus table for one reveals why...
locus_table(pst1_genclone)[1:20,]
# All the stats are missing! Let's just move on I guess?
```

### Phylogenetically uninformative loci
Remove loci that contain less than 2/N % of loci, with N equal to number of individuals in data set
```{r loci}
nLoc(pst1_genclone)
iPst <- informloci(pst1_genclone)
nLoc(iPst)
# Notice there are fewer loci - how does this affect association?
poppr(pst1_genclone)
poppr(iPst)
```
# Analysis
## DAPC
### Geography
First take a look at geography.
```{r DAPC, message=FALSE}
# for genclone
# Scatter plot
dapc.iPst <- dapc(iPst, var.contrib = TRUE, center = FALSE, n.pca = 50,  n.da = nPop(iPst) - 1)
optim.a.score(dapc.iPst)
dapc.iPst <- dapc(iPst, var.contrib = TRUE, center = FALSE, n.pca = 17, n.da = 2)
scatter(dapc.iPst, mstree = TRUE, lwd = 2, lty = 2)

# Compoplot
dapc.results <- as.data.frame(dapc.iPst$posterior)
dapc.results$pop <- pop(iPst)
dapc.results$indNames <- rownames(dapc.results)
dapc.results <- melt(dapc.results)
colnames(dapc.results) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")
p <- ggplot(dapc.results, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
p <- p + geom_bar(stat='identity') 
p <- p + scale_fill_manual(values = rainbow(4)) 
p <- p + facet_grid(~Original_Pop, scales = "free")
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
p

# for genlight
# dapc.iPst_genlight <- dapc(pst1_genlight, var.contrib = TRUE, n.pca = 30,  n.da = nPop(iPst) - 1)
# optim.a.score(dapc.iPst)
# dapc.iPst_genlight <- dapc(pst1_genlight, var.contrib = TRUE, n.pca = 7, n.da = 2)
# scatter(dapc.iPst_genlight, mstree = TRUE, lwd = 2, lty = 2)
```
### Sample source
```{r structure 2, fig.width=12}
# Establish source specifications
setPop(pst1_genclone) <- ~Source
setPop(pst1_genlight) <- ~Source
iPst_source <- informloci(pst1_genclone)
```
```{r sample source continued}
#for genclone
# Scatter plots
dapc.iPst_source <- dapc(iPst_source, var.contrib = TRUE, center = FALSE, n.pca = 25,  n.da = nPop(iPst) - 1)
optim.a.score(dapc.iPst_source)
dapc.iPst_source <- dapc(iPst_source, var.contrib = TRUE, center = FALSE, n.pca = 19, n.da = 2)
scatter(dapc.iPst_source, mstree = TRUE, lwd = 2, lty = 2)

# Compoplots
dapc.results <- as.data.frame(dapc.iPst_source$posterior)
dapc.results$pop <- pop(iPst_source)
dapc.results$indNames <- rownames(dapc.results)
library(reshape2)
dapc.results <- melt(dapc.results)
colnames(dapc.results) <- c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")
p <- ggplot(dapc.results, aes(x=Sample, y=Posterior_membership_probability, fill=Assigned_Pop))
p <- p + geom_bar(stat='identity') 
p <- p + scale_fill_manual(values = rainbow(4)) 
p <- p + facet_grid(~Original_Pop, scales = "free")
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8))
p

# for genlight
# dapc.iPst_source_genlight <- dapc(pst1_genlight, var.contrib = TRUE, n.pca = 30,  n.da = nPop(pst1_genlight) - 1)
# scatter(dapc.iPst_source_genlight, mstree = TRUE, lwd = 2, lty = 2)
```
Looks like there MIGHT be different between source. We'll have to do some more analyses later to find out for sure though. There is a lot of overlap and some things that don't make sense - I would expect root and soil to be more closely linked than root and stem. I also would expect stem and leaf to be more closely linked than root and stem. These are actually in similar environments.

## Trees
We'll use Provesti's distance, also known as bitwise distance. It can use any marker and CAN handle mising data - which we definitely have.
Don't forget that:

genclone object | Population metric | Location
--------------- | ----------------- | ---------
iPst            |Town_County        | Where on the Oregon map
iPst_Source     |Source             | Where on the plant system

```{r genetic distance, fig.height=15, error = TRUE}
# rough tree, not bootstrapped
# dist <- provesti.dist(iPst)
# tree <- dist %>% # using distance just calculated
#   nj() %>% # calculate neighbor-joining tree
#   ladderize() # organize branches by clade
# plot(tree)
# add.scale.bar(length = 0.05) # scale bar showing 5% diff.

# bootstrapped tree
geo_tree <- aboot(iPst, dist = nei.dist, sample = 200, tree = "nj", cutoff = 50, quiet = TRUE)
library("ape")
library(phangorn)
cols <- rainbow(4)
plot.phylo(ladderize(midpoint(geo_tree)), cex = 0.8, font = 2, adj = 0, tip.color = cols[pop(iPst)],
           label.offset = 0.0125, type = "unrooted")
nodelabels(geo_tree$node.label, adj = c(1.3, -0.5), frame = "n", cex = 0.8,
           font = 3, xpd = TRUE)
legend('topleft', legend = levels(pop(iPst)), fill = cols, border = FALSE, bty = "n", cex = 2)
#axisPhylo(3)
aboot(iPst_source, dist = provesti.dist, sample = 200, tree = "nj", cutoff = 50, quiet = TRUE)
```
##Fst
```{r fst, fig.width=16}
# for genclone
fst_iPst <- pairwise.fst(iPst, res.type="matrix")
fst_iPst_source <- pairwise.fst(iPst_source, res.type = "matrix")

fst_iPst
heatmap.2(fst_iPst, dendrogram = "column", "Town_County of origin",
          srtCol = 15, srtRow = 60)
fst_iPst_source
heatmap.2(fst_iPst_source, dendrogram = "column", main = "Sample Source",
          srtCol = 15, srtRow = 60)

# for genlight - DOES NOT WORK
# setPop(pst1_genclone) <- ~Source
# fst_iPst <- pairwise.fst(pst1_genclone, res.type="matrix")
# setPop(pst1_genclone) <- ~Town/County
# fst_iPst_source <- pairwise.fst(pst1_genclone, res.type = "matrix")
# 
# fst_iPst
# heatmap.2(fst_iPst, dendrogram = "column", "Town_County of origin",
#           srtCol = 15, srtRow = 75)
# fst_iPst_source
# heatmap.2(fst_iPst_source, dendrogram = "column", main = "Sample Source",
#           srtCol = 15, srtRow = 75)
```

##PCA
```{r other tools}
# for genlight only - glPca doesnt recognize genclones
# First by geography
setPop(pst1_genlight) <- ~Town/County
setPop(pst1_genclone) <- ~Town/County
iPst.pca <- glPca(pst1_genlight, nf = 3)
barplot(iPst.pca$eig, col = heat.colors(50), main = "PCA Eigenvalues")
iPst.pca.scores <- as.data.frame(iPst.pca$scores)
iPst.pca.scores$pop <- pop(pst1_genclone)
set.seed(9)
p <- ggplot(iPst.pca.scores, aes(x=PC1, y=PC2, colour=pop)) 
p <- p + geom_point(size=2)
p <- p + stat_ellipse(level = 0.95, size = 1)
p <- p + scale_color_manual(values = rainbow(4)) 
p <- p + geom_hline(yintercept = 0) 
p <- p + geom_vline(xintercept = 0) 
p <- p + theme_bw()
p

# Now by source
setPop(pst1_genlight) <- ~Source
setPop(pst1_genclone) <- ~Source
iPst.pca <- glPca(pst1_genlight, nf = 3)
barplot(iPst.pca$eig, col = heat.colors(50), main = "PCA Eigenvalues")
iPst.pca.scores <- as.data.frame(iPst.pca$scores)
iPst.pca.scores$pop <- pop(pst1_genclone)
set.seed(9)
p <- ggplot(iPst.pca.scores, aes(x=PC1, y=PC2, colour=pop)) 
p <- p + geom_point(size=2)
p <- p + stat_ellipse(level = 0.95, size = 1)
p <- p + scale_color_manual(values = rainbow(4)) 
p <- p + geom_hline(yintercept = 0) 
p <- p + geom_vline(xintercept = 0) 
p <- p + theme_bw()
p
```

## Minimum Spanning Network
```{r lib, results="hide", message=FALSE}
library(igraph)
library(RColorBrewer)
```
```{r msn, error=TRUE}
# First geography
Pst.dist <- bitwise.dist(iPst)
Pst.msn <- poppr.msn(iPst, Pst.dist, showplot = FALSE, include.ties = T)

node.size <- rep(2, times = nInd(iPst))
names(node.size) <- indNames(iPst)
vertex.attributes(Pst.msn$graph)$size <- node.size

set.seed(9)
plot_poppr_msn(iPst, Pst.msn, palette=brewer.pal(n = nPop(iPst), name = "Dark2"), gadj = 70)

# Now source
Pst.dist_source <- bitwise.dist(iPst_source)
Pst.msn_source <- poppr.msn(iPst_source, Pst.dist_source, showplot = FALSE, include.ties = T)

node.size <- rep(2, times = nInd(iPst_source))
names(node.size) <- indNames(iPst_source)
vertex.attributes(Pst.msn_source$graph)$size <- node.size

set.seed(9)
plot_poppr_msn(iPst_source, Pst.msn_source, palette=brewer.pal(n = nPop(iPst_source),
                                                                      name = "Dark2"), gadj = 70)
```

## Linkage Disequilbrium
```{r ld}
iPst_county <- iPst
setPop(iPst_county) <- ~County
Clack <- popsub(iPst_county, "Clackamas")
Benton <- popsub(iPst_county, "Benton")
Yam <- popsub(iPst_county, "Yamhill")
#ia(iPst, sample = 99)
ia(Clack, sample = 99)
ia(Benton, sample = 99)
ia(Yam, sample = 99)
```
## AMOVA
```{r amova, message = FALSE, error = FALSE}
iPst_amova_town <- poppr.amova(iPst, ~Town, clonecorrect = TRUE)
iPst_amova_county <- poppr.amova(iPst, ~County, clonecorrect = TRUE)
iPst_amova_source <- poppr.amova(iPst, ~Source, clonecorrect = TRUE)
set.seed(1776)
iPst_town_signif <- randtest(iPst_amova_town, nrepet = 999)
iPst_county_signif <- randtest(iPst_amova_county, nrepet = 999)
iPst_source_signif <- randtest(iPst_amova_source, nrepet = 999)
# Compare to a random structure...
iPst.new <- iPst
set.seed(1901)
strata(iPst.new) <- strata(iPst)[sample(nInd(iPst)), -1]
iPst_town_random <- poppr.amova(iPst.new, ~Town)
iPst_town_random
iPst_town_random_signif <- randtest(iPst_town_random, nrepet = 999)

# now compare... there is definitely something going on here - pval of vetween Town variation is much different.
iPst_town_signif
iPst_county_signif
iPst_town_random_signif
```