---
title: "GBS Raw Reads to PopGen of *Phytophthora Plurivora*"
author: "Nick Carleson"
date: "December 11, 2017"
output:
  html_document:
#    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, results='hide', message=FALSE, warning=FALSE}
library(ggplot2)
library(vcfR)
library(reshape2)
library(cowplot)
library(dplyr)
library(forcats)
```
# Variant Calling

## What the data is
We start this process in the command line, on the CGRB Server. Here are some important considerations for our data:

* Illumina HiSeq 3000
* 150bp read length
* single-end reads
* est. 80-120 Mb _Phytophthora_ genome
* 96 samples with technical reps
* 5-10bp variable barcode length
* PstI-MspI digestion combination
    + SbfI-MspI also used but coverage was worse

For this project, we'll use `Stacks 1.47` for _de novo_ calling of variants. Main references for help are:

* Catchen et al. Stacks: building and genotyping loci de novo from short-read sequences. _G3: Genes, genomes, genetics_ 2011, __3,__ 171-182
* Catchen et al. Stacks: an analysis tool set for population genomics. _Molecular Ecology_ 2013, __11,__ 3124-3140
* Paris et al. Lost in parameter space: a road map for `STACKS`. _Methods in Ecology and Evolution_ 2017, __8,__ 1360–1373

```{r bash, engine = 'bash', eval = FALSE}
cut -d ',' -f 3,4 key_GBS0102_uniquenames_fixedsample.csv | tail -n +2 | sed 's/,/\t/g' > barcode_sorted_unique_fixed.txt
```
the first 10 lines look like this, with barcode on the left and sample name on the right. It's tab separated.

```{r sample head, engine = 'bash', eval = FALSE, highlight = FALSE}
head barcode_sorted_unique_fixed.txt
TGACGCCA	M-04-R
CAGATA	M-16-S_1
TTGCTG	M-16-S_2
GAAGTG	M-22-S3
TAGCGGAT	M-24-R
TATTCGCAT	M-30-R
ATAGAT	M-4311
CGCAACCAGT	M-4313
CCGAACA	M-4314
GGAAGACAT	M-4319_1
```

## Running stacks
And now for the Stacks _de novo_ pipeline! The main step is denovo_map.pl, which is a wrapper script that runs the core commands of stacks. We will run it using the `populations` setting - not the `genotypes` setting; we don't have parents in our population. We will run this analysis with a few different population scenarios - Town, County, and Source. We want to see what the Fst statistics are for each of these. 
```{r bash pipeline, engine = 'bash', eval = FALSE}
# demultiplex, quality filter
process_shortreads -f ../raw_data_links/ALL_AH7HTYBBXX_s_3_fastq.txt.gz -b barcode_sorted_unique_fixed.txt \
-o ~/grunwald_lab_me/plurivora/demultiplexed_process_shortreads_3/ -i gzfastq -q -D
```
```{r bash pipeline2, engine = 'bash', eval = FALSE}
# while that's running, make population maps to analyze
awk -F ',' '{ print $1"\t"$2 }' popmap_stacks_pstI.csv | tail -n +2 > popmap_isolate.tsv
awk -F ',' '{ print $1"\t"$4 }' popmap_stacks_pstI.csv | tail -n +2 > popmap_source.tsv
awk -F ',' '{ print $1"\t"$5 }' popmap_stacks_pstI.csv | tail -n +2 > popmap_nursery.tsv
awk -F ',' '{ print $1"\t"$6 }' popmap_stacks_pstI.csv | tail -n +2 > popmap_town.tsv
awk -F ',' '{ print $1"\t"$7 }' popmap_stacks_pstI.csv | tail -n +2 > popmap_county.tsv
awk -F ',' '{ print $1"\t"$6"\t"$7 }' popmap_stacks_pstI.csv | tail -n +2 > popmap_town_county.tsv
```
```{r bash pipeline3, engine = 'bash', eval = FALSE}
# make all directories necessary
mkdir denovo_map/nursery denovo_map/county denovo_map/isolate denovo_map/source
mkdir denovo_map/town denovo_map/town_county
```
```{r bash pipeline4, engine = 'bash', eval = FALSE}
# automatically run ustacks -> cstacks -> sstacks -> population
# change out town_county below with each type of metadata
# should be an array job, where that changes out depending of row on list you're on
# but I was too lazy. Don't be me. Don't be too lazy
denovo_map.pl -o ./denovo_map/town_county -T 8 -O ../popmap_town_county.tsv --samples ./ \
-X "populations:--fstats" -X "populations:--log_fst_comp" \
-X "populations:--vcf" -X "populations:--plink" -X "populations:--genepop" \
-m 5 -M 5 -n 6 -S -b 1
```
This data set is ~365 million reads... it took stacks a very long time to complete. I wasn't sure if it was running properly, but running on subsets of 1.2m, 2.4m, and 3.6m reads showed the process was running properly. Scripts to complete this subsampling are on github as:

* subsample_fastq.sh
* subsample_fastq_weedlist.sh

full commands (7 steps) and files required to execute successfully are in `pipeline_rerun_denovo_map_subset.txt`.


# VCF Wrangling
let's look at one of the vcf file outputs from a sample run to see if it worked at all
```{r read_vcf, results='hide', message=FALSE, warning=FALSE}
library(vcfR)
vcf <- read.vcfR('~/Data/plurivora/PstI/stacks_121317/town/unfiltered_vcf/batch_1.vcf.gz')
```

## Filtering
Since we called our SNPs *de novo*, the best filtering we can do is based on depth.

Set 0 values in VCF to 'NA'
```{r missing_first}
vcf_0toMissing <- vcf
dp <- extract.gt(vcf_0toMissing, element = "DP", as.numeric = TRUE)
dp[dp == 0] <- NA
vcf_0toMissing@gt[,-1][ is.na(dp) == TRUE] <- NA
```
Before censoring 0's to NA:
`r show(vcf)`
Now, here's how much missing data we start off with:
`r show(vcf_0toMissing)`

### Before filtering
Let's take a look at our depth
```{r dp}
dp <- extract.gt(vcf, element = "DP", as.numeric = TRUE)
plot(rowSums(dp))
```

Looks like there's a huge amount of variants have VERY low counts across all samples.  
What does this look like in our violin plots?
```{r Violin, fig.width=15, fig.height=15}
# First create ggplot-friendly chart
dpf <- melt(dp, varnames=c('Index', 'Sample'), value.name = 'Depth', na.rm=TRUE)
dpf_2 <- dpf[ dpf$Depth > 0,]
# Now make violin plots by sample
samps_per_row <- 14
myRows <- ceiling(length(levels(dpf_2$Sample))/samps_per_row)
myList <- vector(mode = "list", length = myRows)

# Make list of necessary plots
for(i in 1:myRows){
  myIndex <- c(i*samps_per_row - samps_per_row + 1):c(i*samps_per_row)
  myIndex <- myIndex[myIndex <= length(levels(dpf_2$Sample))]
  myLevels <- levels(dpf_2$Sample)[myIndex]
  myRegex <- paste(myLevels, collapse = "$|^")
  myRegex <- paste("^", myRegex, "$", sep = "")
  myList[[i]] <- dpf_2[grep(myRegex, dpf_2$Sample),]
  myList[[i]]$Sample <- factor(myList[[i]]$Sample)
}

# Create the plot.
myPlots <- vector(mode = "list", length = myRows)
for(i in 1:myRows){
  myPlots[[i]] <- ggplot(myList[[i]], aes(x=Sample, y=Depth)) + 
    geom_violin(fill="#8dd3c7", adjust=1.0, scale = "count", trim=TRUE)
  
  myPlots[[i]] <- myPlots[[i]] + theme_bw()
  myPlots[[i]] <- myPlots[[i]] + theme(axis.title.x = element_blank(), 
                                       axis.text.x = element_text(angle = 60, hjust = 1))
  myPlots[[i]] <- myPlots[[i]] + scale_y_continuous(trans=scales::log2_trans(), 
                                                    breaks=c(1, 10, 100, 800),
                                                    minor_breaks=c(1:10, 2:10*10, 2:8*100))
  myPlots[[i]] <- myPlots[[i]] + theme( panel.grid.major.y=element_line(color = "#A9A9A9", size=0.6) )
  myPlots[[i]] <- myPlots[[i]] + theme( panel.grid.minor.y=element_line(color = "#C0C0C0", size=0.2) )
}

# Plot the Plot
# Here's how each restriction enzyme performed, separated by variant
plot_grid(plotlist = myPlots, nrow = myRows)
```
Looks OKAY. Some samples are clearly worse than others.

### During

#### Censorship
Censor all values based on depth - we'll pick a 95% quantile cutoff as too high. We'll also censor any value that has a depth of less than 4.
```{r filter_95}
# re-extract dp just to validate we are doing the right thing...
dp <- extract.gt(vcf, element = "DP", as.numeric = TRUE)
# Calculate quantiles
quants <- apply(dp, MARGIN=2, quantile, probs=c(0.05, 0.95), na.rm=TRUE)

# Sweep up the file to get rid of the quartiles
dp2 <- sweep(dp, MARGIN=2, FUN="-", quants[1,])
dp[dp2 < 0] <- NA
dp2 <- sweep(dp, MARGIN=2, FUN="-", quants[2,])
dp[dp2 > 0] <- NA
vcf_95 <- vcf
vcf_95@gt[,-1][ is.na(dp) == TRUE] <- NA

# Also censor anything with a count of less than 4
dp[dp < 4] <- NA
vcf_95@gt[,-1][ is.na(dp) == TRUE] <- NA
vcf_95
```

#### Omission
Censoring just changes abnormal values to "NA", which have now become missing samples. Omit some samples based on criteria of NA-ness.

Omit *samples* with missingness greater than 95%
```{r omit_samples}
# extract depth again, just to be sure we are on newest VCF version
dp <- extract.gt(vcf_95, element = "DP", as.numeric = TRUE)
# Now omitting samples
myMiss <- apply(dp, MARGIN = 2, function(x){ sum( is.na(x) ) } )
myMiss <- myMiss / nrow(dp)
vcf_95_samples <- vcf_95
vcf_95_samples@gt <- vcf_95_samples@gt[, c(TRUE, myMiss < 0.95)]
vcf_95_samples
```

Omit *variants* with missingness greater than 10%
```{r omit_variants}
# extract depth again...
dp <- extract.gt(vcf_95_samples, element = "DP", as.numeric = TRUE)
# Now omitting vars
myMiss <- apply(dp, MARGIN = 1, function(x) {sum(is.na(x))})
myMiss <- myMiss / ncol(dp)
vcf_95_samples_vars <- vcf_95_samples
vcf_95_samples_vars <- vcf_95_samples_vars[myMiss < 0.1, ]
vcf_95_samples_vars
```

Now let's get variables to describe the new VCF file
```{r final_stats}
# depth matrix after omitting variants and samples
dp <- extract.gt(vcf_95_samples_vars, element = "DP", as.numeric = TRUE)
heatmap.bp(dp[1:1000,], rlabels = FALSE)

# amounts of missing data after omitting variants and samples
myMiss <- apply(dp, MARGIN = 2, function(x){ sum( is.na(x) ) } )
myMiss <- myMiss / nrow(dp)
```

The final heatmap doesn't look good for some samples, now... Let's filter *samples* by missingness again. Based on what the missingness looks like overall, we can decide on a new cutoff
`r myMiss`
50% Looks to be pretty good - there's a clear distinction between samples above this threshold and below it.
```{r samples2}
# Now filter
vcf_95_samples2_vars <- vcf_95_samples_vars
vcf_95_samples2_vars@gt <- vcf_95_samples2_vars@gt[, c(TRUE, myMiss < 0.50)]
vcf_95_samples2_vars
```

Depth plot of final vcf:
```{r final_plot}
dp <- extract.gt(vcf_95_samples2_vars, element = "DP", as.numeric = TRUE)
heatmap.bp(dp[1:1000,], rlabels = FALSE)
```


### After
The evolution of our VCF.
```{r final_vcf}
#Before processing:
vcf
#After changing 0 counts to missing data:
vcf_0toMissing
#After censoring variants (converting them to "missing"") with a depth more than 95% of all other variants in each sample:
vcf_95
#After filtering out samples with more than 95% missing data:
vcf_95_samples
#After filtering out variants with a missingness greater than 10%:
vcf_95_samples_vars
#After then filtering out samples again, if they had greater than 50% missing data, we received our final VCF. Here's what it looks like:
vcf_95_samples2_vars
```
This final VCF looks good! Let's write it to file, export it to a genind (poppr/adegenet/ape) and PED/BED (Admixture)

```{r file_finish}
# VCF to genind
final_genind <- vcfR2genind(vcf_95_samples2_vars)
# VCF to genlight
final_genlight <- vcfR2genlight(vcf_95_samples2_vars)
# Write VCF to file
write.vcf(vcf_95_samples2_vars, file = "~/Data/plurivora/PstI/stacks_121317/filtered_snps.vcf.gz")
# Save VCF, genind, genlight as R objects to load easily
save(vcf_95_samples2_vars, file = "~/Data/plurivora/PstI/stacks_121317/final_vcf")
save(final_genind, file = "~/Data/plurivora/PstI/stacks_121317/final_genind")
save(final_genlight, file = "~/Data/plurivora/PstI/stacks_121317/final_genlight")
```

To convert to `BED`, we'll need to make a metadata file (FAM). Fortunately, stacks generated `FAM` files for us during its `run_denovo_map` workflow. We do need to subset the `fam`s based on samples we ended up retaining
```{r subset_assist}
# make list of samples to use
list <- colnames(vcf_95_samples2_vars@gt)
list_noformat <- list[2:length(list)]

# town fam
subset_fam <- function(x, list, file) {
  fam <- read.table(x)
  #all_samples <- fam$V2
  fam_subset <- subset(fam, V2 %in% list)
  write.table(fam_subset, file = file, sep = "\t",
              col.names = FALSE, row.names = FALSE,
              quote = FALSE)
}

# town fam
subset_fam(x = "~/Data/plurivora/PstI/stacks_121317/town/unfiltered_vcf/batch_2.fam",
           list = list_noformat, file = "~/Data/plurivora/PstI/stacks_121317/subset_town_batch_2.fam")
# county fam
subset_fam(x = "~/Data/plurivora/PstI/stacks_121317/county/unfiltered_vcf/batch_2.fam",
           list = list_noformat, file = "~/Data/plurivora/PstI/stacks_121317/subset_county_batch_2.fam")
```
Inspecting the fam files in bash looks okay! Now let's do the conversion.
This requires `GATK`

```{r bash_conv, engine='bash', eval=FALSE}
cd ~/Data/plurivora/PstI/stacks_121317
gunzip filtered_snps.vcf.gz
vcftools --vcf filtered_snps.vcf --plink --out filtered_plink
```
# Admixture
None of this is probably legit because I haven't done filtering for coverage yet. I probably should use vcfR (below) to filter and then create the bed.

## File prep
First, Admixture. Admixture didn't actually work using PED files, I had to convert tob BED first :/
```{r file_prep, engine='bash', eval=FALSE}
# Convert to binary files
plink --file filtered_plink --maf 0.05 --make-bed --out filtered_plink_bin --noweb
# Notice the fam file created doesn't contain any useful population associations
# We'll use the fam files created via subsetting previously ( subset_fam(x, list, file) )
cp filtered_plink_bin.bed filtered_plink_bin_county.bed
cp filtered_plink_bin.bim filtered_plink_bin_county.bim
mv filtered_plink_bin.bed filtered_plink_bin_town.bed
mv filtered_plink_bin.bim filtered_plink_bin_town.bim
mv subset_county_batch_2.fam filtered_plink_bin_county.fam
mv subset_town_batch_2.fam filtered_plink_bin_town.fam
```
## Analyis
### Cross-validation
```{r admixture, engine = 'bash', eval = FALSE, results='hide', message=FALSE, warning=FALSE}
# Run cross-validation
for K in 1 2 3 4 5 6 7 8 9 10 11 12; do admixture \
--cv filtered_plink_bin_town.bed $K | tee town_log${K}.out; done
for K in 1 2 3 4 5 6 7 8 9 10 11 12; do admixture \
--cv filtered_plink_bin_county.bed $K | tee county_log${K}.out; done

# Write Cross-Validation values to table
# cross_validation_tabler.sh is on GitHub here:
# https://github.com/Neato-Nick/variant-calling/blob/master/cross_validation_tabler.sh
~/Scripts/cross_validation_tabler.sh town_log*.out
~/Scripts/cross_validation_tabler.sh county_log*.out
```

Get that into an R data-frame. The cross-validation results should be the same in both population runs. So, we'll only inspect one cross-validation plot. But, the --cv option just additionally runs cross-validation on top of the normal admixture run.
```{r cv}
cv_town <- read.table("~/Data/plurivora/PstI/stacks_121317/town/cv.txt")
cv_town <- cv_town[order(cv_town$V1),]
rownames(cv_town) <- cv_town$K
colnames(cv_town) <- c("K", "CV")
cv_town
ggplot(cv_town[2:length(cv_town$K),], mapping = aes(y = CV, x = K)) +
  geom_line(alpha = 1.0) +
  geom_point(alpha = 1.0) +
  ylab("cross-validation error") +
  scale_x_continuous(breaks = c(2, 4, 6, 8, 10, 12))
```
There are multiple number of clusters that have low error...lets see all the structure plots

### Structure plots
Let's see what the "Structure" plot looks like for all of these cross-validation points
```{r structure plot, fig.width=12}
plot_Q <- function(tbl, fam) {
  # assign sample names to numbers
  tbl$Sample <- fam$V2
  tbl$PopName <- fam$V1

  
  tbl2 <- melt(tbl, id = c("Sample", "PopName"))

  colnames(tbl2) <- c("Sample", "PopName", "Cluster", "Composition")

  # ddply solution
  tbl3 <- tbl2 %>%
    group_by(PopName, Cluster) %>%
    summarise(avg = mean(Composition)) %>%
    right_join(tbl2, by = c("PopName", "Cluster")) %>%
    arrange(PopName, desc(avg), desc(Composition)) %>%
    select(-avg) %>%
    ungroup
  #show(tbl3)
  # Make my own greyscale function...
  #cluster_count <- length(unique(tbl3$Cluster))
  #bw_cols=colorRampPalette(colors=c('#000000', '#ffffff'))(cluster_count)
  ggplot(tbl3) +
    geom_bar(mapping = aes(x = fct_inorder(Sample), y = Composition, fill = Cluster),
             stat = "identity") +
    ylab("% Assignment") +
    xlab("Sample") +
    # for making own greyscale theme. It didn't work.
    #scale_colour_manual(values = c('#000000', '#FFFFFF'))
    #scale_fill_grey() +
    facet_grid(. ~ PopName, scales = "free_x") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) 

}
# fam file contains metadata
fam1 <- read.table("~/Data/plurivora/PstI/stacks_121317/town/filtered_plink_bin_town.fam")

plot_Q(read.table("~/Data/plurivora/PstI/stacks_121317/town/filtered_plink_bin_town.3.Q"),
       fam1)
plot_Q(read.table("~/Data/plurivora/PstI/stacks_121317/town/filtered_plink_bin_town.4.Q",
                  stringsAsFactors = FALSE),
       fam1)
plot_Q(read.table("~/Data/plurivora/PstI/stacks_121317/town/filtered_plink_bin_town.5.Q",
                  stringsAsFactors = FALSE),
       fam1)
plot_Q(read.table("~/Data/plurivora/PstI/stacks_121317/town/filtered_plink_bin_town.6.Q"),
       fam1)
plot_Q(read.table("~/Data/plurivora/PstI/stacks_121317/town/filtered_plink_bin_town.7.Q"),
       fam1)
plot_Q(read.table("~/Data/plurivora/PstI/stacks_121317/town/filtered_plink_bin_town.8.Q"),
       fam1)
```
